---
title: "Differential Expression Results"
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h

output:
  pdf_document:
    fig_caption: true
    keep_tex: true
    latex_engine: pdflatex
  html_document:
    df_print: paged
urlcolor: blue
---

```
{r echo=FALSE, message=FALSE}

# should add a check for this
source("deseq2_config.R")
```

```{r setup3, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.extra = '')
#knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_knit$set(root.dir = rootdir)
```

# DESeq2 Summary Report (with BWA mapping) {#deseq2}

This analysis uses DESeq2 to define differentially regulated genes across conditions. The first section deals
with DESeq2 quality assessments and the later sections with differentially expressed genes.

## Number of genes detected

This section provides a look at the number of genes detected i.e read count >=1

```{r detected-genes, fig.cap='Number of genes detected in each sample', fig.height=2.5, fig.width=5, echo=FALSE, message=FALSE}

# get helper functions
source(helper)

# knitr for tables
library(knitr)

mat <- read.csv(feature.counts, header=T, stringsAsFactors=F, sep="\t", row.names=1)
ngenes <- data.frame(colSums(mat >= 1))

# make conditions
ngenes$cond <- makeConds(rownames(ngenes))
colnames(ngenes) <- c("ngenes", "cond")

# plot number of genes
ggplot(ngenes, aes(x=cond, y=ngenes, colour=cond)) + geom_boxplot() + geom_jitter() + theme_bw()

```

### ANOVA of number of genes detected

This is a test of significant differences in the number of genes detected.

```{r sig-difference-in-genes-detected}

summary(aov(ngenes$ngenes~ngenes$cond))

```

\newpage

## Quality control of DESeq2 normalisation procedures

```{r diff-expression, fig.cap='Mean-variance and dispersion estimates from DESeq2', fig.height=4, fig.width=8,  echo=FALSE, message=FALSE}

library(DESeq2)
library(vsn)
library(RSQLite)
library(grid)
library(gridBase)

####################
# analysis
####################

mat <- read.csv("genes.tsv.gz", header=T, stringsAsFactors=F, sep="\t", row.names=1)
coldata <- makeColData(mat)

dds <- DESeqDataSetFromMatrix(countData = mat,
                              colData = coldata,
			      design = ~ condition)

# add features (gene names)
featureData <- makeFeatureData(annotations.db, mat)
mcols(dds) <- DataFrame(mcols(dds), featureData)

# filtering > a in at least k samples (i.e. all of one condition)
keep <- rowSums(counts(dds) >= a) >= k
dds <- dds[keep,]

# make sure conditions are factors
dds$condition <- factor(dds$condition)

# run analysis
dds.lrt <- DESeq(dds, test="LRT", reduced=~1)
res <- results(dds.lrt)

# read in rlog data NB: SHOULD HAVE USED SAME FILTERING CRITERIA AS FOR
# ANALYSIS. The rlog file is not produced here.
rld <- read.csv(norm.file, sep="\t", stringsAsFactors=F, header=T)
rownames(rld) <- rld$gene_id
rld <- rld[,c(1:ncol(rld)-1)]

# plot mean-variance relationship and dispersion
# estimates

par(mfrow=c(1,2))
plotMeanSd(rld)
plotDispEsts(dds.lrt)

```

\newpage

# Principle components analysis

Here we perform principle components analysis on normalised read counts (rlog) from DESeq2. This gives an
idea of the structure of the data and whether there is an obvious difference between groups.

```{r pca, fig.cap='Principle components analysis', fig.height=15, fig.width=20,  echo=FALSE, message=FALSE}

df <- rld
pc <- PCA(df)
pcs <- getPCA(pc)

# plot cumulative proportion
p1 <- plotCumulativeProportion(pc)

# plot principle components
p2 <- plotPCA(pc, pcs=c("PC1", "PC2"))
p3 <- plotPCA(pc, pcs=c("PC1", "PC3"))
p4 <- plotPCA(pc, pcs=c("PC2", "PC3"))

layout = rbind(c(1,1,1), c(2,3,4))
grid.arrange(grobs=list(p1,p2,p3,p4), layout_matrix=layout)

```

# Overall differential expression analysis

Here we describe genes that were called as differentially expressed. If there are multiple levels per factor
then it is an overall test (ANOVA-like) whereas if just two conditions are present then it is the direct contrast.

```{r de, fig.cap='Differential expression analysis', fig.height=15, fig.width=10,  echo=FALSE, message=FALSE}

library(gplots)

res2 <- getResultsTable(res, featureData)

# subset for genes that are differentially regulated
res.diff <- res2[res2$padj < 0.05 & !(is.na(res2$padj)) & !(is.na(res2$log2FoldChange)),]

# get number different
ndiff <- nrow(res.diff)
df <- data.frame("Number DE" = ndiff)
kable(df, caption="number of genes differentially expressed at p < 0.05")

# get differentially expressed gene ids
diffgenes <- res.diff$gene_id
diff <- rld[diffgenes, ]

# heatmap differences
heatmapMatrix(diff)

# table the top 50
toshow <- res.diff[order(res.diff$padj, decreasing=F),][1:50,]
kable(toshow, caption="top 50 differentially expressed genes by adjusted pvalue", row.names=F)

```

# Pattern of regulation in the top genes

As the analysis was done ANOVA style we don't clearly know (without specifying the contrasts) the pattern of regulation
of these genes. Here we plot the top (20) genes to see what the pattern is.

```{r goi, fig.cap='Expression pattern in top 20 genes', fig.height=15, fig.width=15,  echo=FALSE, message=FALSE}

# get gene names
rld2 <- rld

# get rid of any weird duplicates
res.diff <- res.diff[!(duplicated(res.diff$gene_name)),]
res.diff <- res.diff[order(res.diff$padj, decreasing=F),]

# make gene names the rownames
rld2 <- rld2[rownames(res.diff),]
rownames(rld2) <- res.diff[rownames(rld2),]$gene_name

# collect grobs for figure
grob.list <- list()
for (i in 1:20){
    goi <- rownames(rld2)[i]
    p <- plotGOI(rld2, goi=goi)
    grob.list[[i]] <- p
    }

grid.arrange(grobs=grob.list, ncol=4, nrow=5)

```
\newpage

# Specific contrasts

We have three different comparisons to look at - PSC vs. healthy, UC vs. healthy and PSC vs. UC. Here we plot
the differentially expressed genes in each contrast. 

```{r contrasts, fig.cap='MA plot for each contrast', fig.height=5, fig.width=15,  echo=FALSE, message=FALSE}

###########################################################
###########################################################
# In order to get specific contrast p-values we need
# to run the wald test for each contrast
# However we only consider things significant if they
# are also significant in the full LRT test
###########################################################
###########################################################

# get contrasts (PSC vs. UC)

dds.psc.uc <- dds
dds.psc.uc <- DESeq(dds.psc.uc, test="Wald")
psc.uc <- getContrast(dds.psc.uc, factor=c("condition"),contrast=c("PSC", "UC"))

# PSC vs. healthy
dds.psc.healthy <- dds
dds.psc.healthy <- DESeq(dds.psc.healthy, test="Wald")
psc.healthy <- getContrast(dds.psc.healthy, factor=c("condition"), contrast=c("PSC", "HEALTHY"))

# UC vs. healthy
dds.uc.healthy <- dds
dds.uc.healthy <- DESeq(dds.uc.healthy, test="Wald")
uc.healthy <- getContrast(dds.uc.healthy, factor=c("condition"), contrast=c("UC", "HEALTHY"))


# full set from LRT
lrt.set <- diffgenes

p1 <- MAPlot(uc.healthy, test.in=T, test.set=lrt.set, title="UC vs. HEALTHY")
p2 <- MAPlot(psc.healthy, test.in=T, test.set=lrt.set, title="PSC vs. HEALTHY")
p3 <- MAPlot(psc.uc, test.in=T, test.set=lrt.set, title="PSC vs. UC")

grid.arrange(p1,p2,p3, nrow=1, ncol=3)

# display tables of the top (50) differentially expressed genes in each contrast
uc.healthy.diff <- getResultsTable(uc.healthy, featureData)
uc.healthy.diff <- uc.healthy.diff[uc.healthy.diff$padj < 0.05 & abs(uc.healthy.diff$log2FoldChange) > lfc & !(is.na(uc.healthy.diff$padj)) & !(is.na(uc.healthy.diff$log2FoldChange)) & uc.healthy.diff$gene_id %in% lrt.set, ]
uc.healthy.diff <- uc.healthy.diff[order(uc.healthy.diff$padj, decreasing=F),]

psc.healthy.diff <- getResultsTable(psc.healthy, featureData)
psc.healthy.diff <- psc.healthy.diff[psc.healthy.diff$padj < 0.05 & abs(psc.healthy.diff$log2FoldChange) > lfc & !(is.na(psc.healthy.diff$padj)) & !(is.na(psc.healthy.diff$log2FoldChange)) & psc.healthy.diff$gene_id %in% lrt.set, ]
psc.healthy.diff <- psc.healthy.diff[order(psc.healthy.diff$padj, decreasing=F),]

psc.uc.diff <- getResultsTable(psc.uc, featureData)
psc.uc.diff <- psc.uc.diff[psc.uc.diff$padj < 0.05 & abs(psc.uc.diff$log2FoldChange) > lfc & !(is.na(psc.uc.diff$padj)) & !(is.na(psc.uc.diff$log2FoldChange)) & psc.uc.diff$gene_id %in% lrt.set, ]
psc.uc.diff <- psc.uc.diff[order(psc.uc.diff$padj, decreasing=F),]

kable(uc.healthy.diff[1:14,], caption="UC vs. HEALTHY (top 14)", row.names=F)
kable(psc.healthy.diff[1:50,], caption="PSC vs. HEALTHY (top 50)", row.names=F)
kable(psc.uc.diff[1:50,], caption="PSC vs. UC (top 50)", row.names=F)

```

\newpage

## Venn diagram of the overlap between contrasts

```{r venn-overlap, fig.cap="Venn of the overlap between contrasts", fig.height=2.5, fig.width=2.5, echo=FALSE, message=FALSE}

library(VennDiagram)

nuc <- nrow(uc.healthy.diff)
npsc <- nrow(psc.healthy.diff)
npsc.uc <- nrow(psc.uc.diff)

noverlap1 <- length(intersect(rownames(uc.healthy.diff), rownames(psc.healthy.diff)))
noverlap2 <- length(intersect(rownames(psc.healthy.diff), rownames(psc.uc.diff)))
noverlap3 <- length(intersect(rownames(uc.healthy.diff), rownames(psc.uc.diff)))

# all
noverlap4 <- length(intersect(intersect(rownames(uc.healthy.diff), rownames(psc.healthy.diff)), rownames(psc.uc.diff)))

categories <- c("UC vs. HEALTHY", "PSC vs. HEALTHY", "PSC vs. UC")

venn1 <- draw.triple.venn(nuc, npsc, npsc.uc, noverlap1, noverlap2, noverlap3, noverlap4, category=categories, fill=c("red", "green", "blue"), euler.d=T, scaled=T, font.family=rep("sans", 7), cat.cex=rep(0.5,3))
grid.draw(venn1)

```

The venn diagram shows us that any (well almost) gene that is regulated in UC is also regulated in PSC i.e. the common aspects
of UC and PSC UC. However there is a substantial component that is specific to PSC. For downstream analysis I will focus
on two sets of genes - those that are regulated (up and down) in UC vs. HEALTHY (this also encompasses a significant proportion
of the differences seen in PSC vs. healthy) and those genes that are specifically regulated in PSC i.e. the genes that
are regulated in PSC vs. healthy and PSC vs. UC.

```{r write-lists, echo=FALSE, message=FALSE}

# lists of UC regulated genes
ucgenes.up <- uc.healthy.diff$gene_name[uc.healthy.diff$log2FoldChange > 0]
ucgenes.dn <- uc.healthy.diff$gene_name[uc.healthy.diff$log2FoldChange < 0]

# lists of PSC regulated genes
psc.healthy.up <- psc.healthy.diff[psc.healthy.diff$log2FoldChange > 0,]
psc.uc.up <- psc.uc.diff[psc.uc.diff$log2FoldChange > 0,]

psc.healthy.dn <- psc.healthy.diff[psc.healthy.diff$log2FoldChange < 0,]
psc.uc.dn <- psc.uc.diff[psc.uc.diff$log2FoldChange < 0,]

pscgenes.up <- intersect(psc.healthy.up$gene_name, psc.uc.up$gene_name)
pscgenes.dn <- intersect(psc.healthy.dn$gene_name, psc.uc.dn$gene_name)

# take out those regulated in UC
pscgenes.up <- setdiff(pscgenes.up, ucgenes.up)
pscgenes.dn <- setdiff(pscgenes.dn, ucgenes.dn)

# write out lists
write.table(ucgenes.up, file="../data/uc_upregulated.tsv", sep="\t", col.names=F, row.names=F, quote=F)
write.table(ucgenes.dn, file="../data/uc_downregulated.tsv", sep="\t", col.names=F, row.names=F, quote=F)

write.table(pscgenes.up, file="../data/psc_upregulated.tsv", sep="\t", col.names=F, row.names=F, quote=F)
write.table(pscgenes.dn, file="../data/psc_downregulated.tsv", sep="\t", col.names=F, row.names=F, quote=F)

# write out background gene list
write.table(res2$gene_name, file="../data/background.tsv", sep="\t", col.names=F, row.names=F, quote=F)

```
\newpage



