---
title: "Pathways analysis"
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h

output:
  pdf_document:
    fig_caption: true
    keep_tex: true
    latex_engine: pdflatex
  html_document:
    df_print: paged
urlcolor: blue
---

```{r setup.pathways, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.extra = '')
#knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_knit$set(root.dir = rootdir)
```

# Pathways and transcription factor motif analysis

I have take the gene lists as described in the last section (UC-regulated and PSC-regulated) and run them against
a variety of pathways databases from the MsigDB: HALLMARK genesets, KEGG genesets, TF motif genesets. There was very
little in the way of enriched pathways among these genes.

## PSC up-regulated genes

### HALLMARK gene sets

No significant sets.

### KEGG gene sets

No significant sets.

### TF motif sets

A number of transcription factor motifs were enriched amongst up-regulated genes in PSC. And are described
below.

```{r pathways, fig.cap="Significantly enriched transcription factor motifs in up-regulated genes in PSC", fig.height=5, fig.width=15, echo=FALSE, message=FALSE}

source(pathways.helper)

pathway.table <- read.csv("/gfs/work/nilott/proj018/analysis/biopsies/enrichment_psc_upregulated/tfs/c3.all.v5.1.symbols.foreground.results", header=T, stringsAsFactors=F, sep="\t")
plotPathways(pathway.table)

```

\newpage

# Shared genes across transcription factor motif sets

It is of interest to see which genes are represented in the transcription factor motifs and which ones are shared between different motifs. Here we build a matrix
of genes that are differentially regulated by PSC (upregulated) and see which ones are predicted to be regulated by the different transcription factors.

For visualisation purposes (and maybe biological ones) I filter out "Unknown" motifs and anything less than 2 fold enriched.

```{r shared-motifs, fig.cap="Differentially regulated genes and predicted transcription factor binding", fig.height=25, fig.width=10, echo=FALSE, message=FALSE}

# get pathways results
pathway.table <- read.csv("/gfs/work/nilott/proj018/analysis/biopsies/enrichment_psc_upregulated/tfs/c3.all.v5.1.symbols.foreground.results", header=T, stringsAsFactors=F, sep="\t")

# take out unknown tf motifs (unnamed)
pathway.table <- pathway.table[grep("UNKNOWN", pathway.table$goid, invert=T),]

# filter on fold enrichment
pathway.table <- pathway.table[pathway.table$ratio >= 2, ]


# get genesets
genesets <- read.csv("/gfs/work/nilott/proj018/analysis/biopsies/enrichment_psc_upregulated/c3.all.v5.1.symbols.tsv", header=F, stringsAsFactors=F, sep="\t")

# get diff genes
genes <- read.csv("psc_upregulated.tsv", header=F, stringsAsFactors=F, sep="\t")

# just get geneset genes (in sig pathways) that are regulated by PSC
total <- buildGenesetGenelist(pathway.table, genesets)
total <- intersect(total, genes$V1)

# build the matrix
mat <- buildGeneMatrix(total, pathway.table, genesets)

# draw heatmap
sepr <- seq(1, nrow(mat), 1)
sepc <- seq(1, ncol(mat), 1)
cols <- colorRampPalette(c("white", "blue"))(2)
distfun=function(x) dist(x, method="binary")
heatmap.2(as.matrix(mat), trace="none", colsep=sepc, rowsep=sepr, sepcolor="grey", margins=c(15,15), col=cols, distfun=distfun)

```

\newpage

### MEF2 expression

We saw an overrepresentation of genes with MEF2 binding motifs (and others) among those that are upregulated by PSC. The MEF2
effect is of particular interest as MEF2A itself is upregulated in PSC (see below).

```{r mef2a, fig.cap="MEF2A expression in PSC vs. UC and Healthy biopsies", fig.height=2.5, fig.width=3.5, echo=FALSE, message=FALSE}

plotGOI(rld2, goi="MEF2A")

```

\newpage

## MEF2 targets

Below are plots of the expression profiles of MEF2 targets predicted from the transcription factor motif analysis. These are upregulated
in PSC vs. UC (and vs. healthy).

```{r mef2targets, fig.cap="Expression of MEF2 target genes", fig.height=15, fig.width=15, echo=FALSE, message=FALSE}

# get all mef2 targets
mef.targets <- getGenesetGenes(genesets, "V$MEF2_Q6_01")

# intersect with psc upregulated genes
mef.targets.up <- intersect(genes$V1, mef.targets)

# plot them
# collect grobs for figure
grob.list <- list()
for (i in 1:length(mef.targets.up)){
    goi <- mef.targets.up[i]
    p <- plotGOI(rld2, goi=goi)
    grob.list[[i]] <- p
    }

grid.arrange(grobs=grob.list, ncol=4, nrow=5)
```

